Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2020-04-27 thirdPaper/draft/typeConeAssociahedra.tex
win one more page...
bold letters third paper

% Cluster algebras
\newcommand{\Trop}[1]{\mathbb{P}_{#1}} % tropical semifield
\newcommand{\positiveExponents}[1]{\left\{#1\right\}_+} % positive exponents in tropical semifield
\newcommand{\negativeExponents}[1]{\left\{#1\right\}_-} % negative exponents in tropical semifield
\newcommand{\seed}{\Sigma} % a cluster
\newcommand{\cluster}{\mathrm{X}} % a cluster
\newcommandx{\clusters}{\mathcal{X}} % all clusters
\newcommand{\coefficients}{\mathrm{P}} % a coefficients tuple
\newcommandx{\variables}[2][1=\B_\circ, 2 = \coefficients_\circ]{\mathcal{V}(#1,#2)} % all variables
\newcommandx{\clusterAlgebra}[2][1=\B_\circ, 2=\coefficients_\circ]{\mathcal{A}\!\left(#1,#2\right)} % cluster algebra
\newcommandx{\principalClusterAlgebra}[1][1=\B_\circ]{\mathcal{A}\left(#1\right)} % principal coefficients cluster algebra
\newcommandx{\principalVariables}[1][1=\B_\circ]{\mathcal{V}(#1)} % all variables, principal case
\newcommandx{\coefficientFreeClusterAlgebra}[1][1=\B_\circ]{\mathcal{A}_{\mathrm{fr}}(#1)} % coefficient free cluster algebra
\newcommandx{\clusterComplex}[1][1=\B_\circ]{\Delta(#1)} % cluster complex
\newcommand{\B}{\mathrm{B}} % b-matrix
\newcommand{\A}[1]{\mathrm{A}({#1})} % b-matrix
\newcommand{\D}{\mathrm{D}} % symmetrizer
\newcommand{\simpleRoot}{\alpha} % simple root
\newcommand{\fundamentalWeight}{\omega} % fundamental weights
\newcommandx{\meshes}[1][1=\B_\circ]{\mathcal{M}(#1)} % meshes

@@

\begin{proposition}
\label{prop:characterizationPolytopalFan}
Let~$\Fan$ be an essential complete simplicial fan in~$\R^n$. Then the following are equivalent for any height vector~$\b{h} \in \R^N$:
\begin{enumerate}
\item The fan~$\Fan$ is the normal fan of the polytope~$P_\b{h} \eqdef \set{\b{x} \in \R^n}{\b{G}\b{x} \le \b{h}}$.
\item For any two adjacent maximal cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ with~$\rays \ssm \{\ray\} = \rays' \ssm \{\ray'\}$,
\[
\alpha \, h_{\ray} + \alpha' \, h_{\ray'} + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, h_{\ray[s]} > 0,
\]
where
\[
\alpha \, \ray + \alpha' \, \ray' + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \ray[s] = 0
\]
is the unique (up to rescaling) linear dependence with~$\alpha, \alpha' > 0$ between the rays of~$\rays \cup \rays'$.
\end{enumerate}
\end{proposition}

@@

If $P = Q + R$, then we say that~$R$ is the \defn{Minkowski difference} of~$P$ and~$Q$, denoted~${R = P - Q}$. Note that this is only defined when~$Q$ is a Minkowski summand of~$P$.
For instance, we have ${P_\b{h} - P_{\b{h}'} = P_{\b{h} - \b{h'}}}$ when $P_{\b{h}'}$ is a Minkowski summand of $P_\b{h}$.
There are two natural ways this construction generalizes to arbitrary pairs of polyhedra.
On the one hand, one can consider that ${P - Q = \set{x \in \R^n}{x + Q \subseteq P}}$.
On the other hand, considering arbitrary differences of support functions gives rise to \defn{virtual polyhedra}, which is the Grothendieck group of the semigroup of polytopes with Minkowski sums~\cite{PukhlikovKhovanskii}.

@@

\begin{theorem}
\label{thm:associahedronLoday}
The associahedron~$\Asso[n]$ can be described equivalently as:
\begin{itemize}
\item the convex hull of the points~$L(T) \in \R^{n+1}$ for all rooted binary trees~$T$ with~$n+1$ internal nodes, where the $i$th coordinate of~$L(T)$ is the product of the number of leaves in the left subtree by the number of leaves in the right subtree of the $i$th node of~$T$ in inorder~\cite{Loday},
\item the intersection of the hyperplane~$\bigset{\b{x} \in \R^{n+1}}{\sum_{\ell \in [n+1]} x_\ell = \binom{n+2}{2}}$ with the halfspaces $\bigset{\b{x} \in \R^{n+1}}{\sum_{a < \ell < b} x_\ell \ge \binom{b-a}{2}}$ for all internal diagonals~$(a,b)$ of \mbox{the $(n+3)$-gon~\cite{ShniderSternberg}},
\item the Minkowski sum of the faces~$\simplex_{[i,j]}$ of the standard $n$-dimensional simplex~$\simplex_{[n+1]}$ corresponding to all proper intervals~$[i,j]$ of~$[n+1]$,~\cite{Postnikov}.
\end{itemize}
\end{theorem}

\enlargethispage{.1cm}
We now focus on the normal fan~$\gvectorFan[n]$ of~$\Asso[n]$.
Since~$\Asso[n]$ lives in a hyperplane of~$\R^{n+1}$, so does its normal fan.
Let~$\HH \eqdef \bigset{\b{x} \in \R^{n+1}}{\sum_{\ell \in [n+1]} x_\ell = 0}$ and~$\pi : \R^{n+1} \to \HH$ denote the orthogonal projection.
The following fans are illustrated in \cref{fig:lodayFans} in dimension~$2$ and~$3$.

\begin{theorem}
\label{thm:normalFanLoday}
In the normal fan~$\gvectorFan[n]$ of~$\Asso[n]$, 
\begin{itemize}
\item the normal vector of the facet corresponding to an internal diagonal~$(a,b)$ of the $(n+3)$-gon is the vector~$\gvector{a,b} \eqdef \pi\big( \sum_{a < \ell < b} \b{e}_\ell \big) = (n+1) \sum_{a < \ell < b} \b{e}_\ell - (b-a-1) \sum_{1 \le \ell \le n+1} \b{e}_\ell$.
\item the normal cone of the vertex corresponding to a rooted binary tree~$T$ is the incidence cone~$\bigset{\b{x} \in \HH}{x_i \le x_j \text{ for all edges $i \to j$ in $T$}}$.
\end{itemize}
\end{theorem}

@@

\para{Principal coefficients and $\b{g}$- and $\b{c}$-vectors}
%
We now consider principal coefficients to define the $\b{g}$- and $\b{c}$-vectors.
As defined in~\cite[Def.~3.1]{FominZelevinsky-ClusterAlgebrasIV}, the cluster algebra with \defn{principal coefficients} at~$\B_\circ$ is the cluster algebra~$\clusterAlgebra[\B_\circ][\coefficients_\circ]$ in~$\Q(x_1, \dots, x_n, p_1, \dots, p_n)$, where~$\coefficients_\circ = \{p_x\}_{x \in \cluster_\circ}$ are precisely the generators~$p_1, \dots, p_n$ of~$\Trop{n}$ relabeled by~$\cluster_\circ$.
For simplicity, we drop the mention of the coefficients for principal cluster algebras in the notations for seeds~$\seed = (\B, \cluster)$, for cluster algebras~$\principalClusterAlgebra$ and for variables~$\principalVariables$.
Principal coefficients cluster algebras are $\Z^n$-graded (in the fundamental weight basis~$\{\omega_x\}_{x \in \cluster_\circ}$ of~$V$) for the degree function~$\deg(\B_\circ,\cdot)$ on $\principalClusterAlgebra$ obtained by setting
\(
\deg(\B_\circ,  x) \eqdef \fundamentalWeight_x
\)
and
\(
\deg(\B_\circ, p_x) \eqdef \sum_{y \in \cluster_\circ} -b_{yx} \fundamentalWeight_y
\)
for any $x\in\cluster_\circ$.

@@

\newpage
\part{Grothendieck group and relations between $\b{g}$-vectors}
\label{part:algebra}

@@

\newpage
\bibliographystyle{alpha}
\bibliography{typeConeAssociahedra}
\label{sec:biblio}

>>>>

% Cluster algebras
\newcommand{\Trop}[1]{\mathbb{P}_{#1}} % tropical semifield
\newcommand{\positiveExponents}[1]{\left\{#1\right\}_+} % positive exponents in tropical semifield
\newcommand{\negativeExponents}[1]{\left\{#1\right\}_-} % negative exponents in tropical semifield
\newcommand{\seed}{\Sigma} % a cluster
\newcommand{\cluster}{\mathrm{X}} % a cluster
\newcommandx{\clusters}{\mathcal{X}} % all clusters
\newcommand{\coefficients}{\mathrm{P}} % a coefficients tuple
\newcommandx{\variables}[2][1=\B_\circ, 2 = \coefficients_\circ]{\mathcal{V}(#1,#2)} % all variables
\newcommandx{\clusterAlgebra}[2][1=\B_\circ, 2=\coefficients_\circ]{\mathcal{A}\!\left(#1,#2\right)} % cluster algebra
\newcommandx{\principalClusterAlgebra}[1][1=\B_\circ]{\mathcal{A}\left(#1\right)} % principal coefficients cluster algebra
\newcommandx{\principalVariables}[1][1=\B_\circ]{\mathcal{V}(#1)} % all variables, principal case
\newcommandx{\coefficientFreeClusterAlgebra}[1][1=\B_\circ]{\mathcal{A}_{\mathrm{fr}}(#1)} % coefficient free cluster algebra
\newcommandx{\clusterComplex}[1][1=\B_\circ]{\Delta(#1)} % cluster complex
\newcommand{\B}{\mathrm{B}} % b-matrix
\newcommand{\A}[1]{\mathrm{A}({#1})} % b-matrix
\newcommand{\D}{\mathrm{D}} % symmetrizer
\newcommand{\simpleRoot}{\b{\alpha}} % simple root
\newcommand{\fundamentalWeight}{\b{\omega}} % fundamental weights
\newcommandx{\meshes}[1][1=\B_\circ]{\mathcal{M}(#1)} % meshes

@@

\begin{proposition}
\label{prop:characterizationPolytopalFan}
Let~$\Fan$ be an essential complete simplicial fan in~$\R^n$. Then the following are equivalent for any height vector~$\b{h} \in \R^N$:
\begin{enumerate}
\item The fan~$\Fan$ is the normal fan of the polytope~$P_\b{h} \eqdef \set{\b{x} \in \R^n}{\b{G}\b{x} \le \b{h}}$.
\item For any two adjacent maximal cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ with~$\rays \ssm \{\ray\} = \rays' \ssm \{\ray'\}$,
\[
\alpha \, \b{h}_{\ray} + \alpha' \, \b{h}_{\ray'} + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \b{h}_{\ray[s]} > 0,
\]
where
\[
\alpha \, \ray + \alpha' \, \ray' + \sum_{\ray[s] \in \rays \cap \rays'} \beta_{\ray[s]} \, \ray[s] = 0
\]
is the unique (up to rescaling) linear dependence with~$\alpha, \alpha' > 0$ between the rays of~$\rays \cup \rays'$.
\end{enumerate}
\end{proposition}

@@

If $P = Q + R$, then we say that~$R$ is the \defn{Minkowski difference} of~$P$ and~$Q$, denoted~${R = P - Q}$. Note that this is only defined when~$Q$ is a Minkowski summand of~$P$.
For instance, we have ${P_\b{h} - P_{\b{h}'} = P_{\b{h} - \b{h'}}}$ when $P_{\b{h}'}$ is a Minkowski summand of $P_\b{h}$.
There are two natural ways this construction generalizes to arbitrary pairs of polyhedra.
On the one hand, one can consider that ${P - Q = \set{\b{x} \in \R^n}{\b{x} + Q \subseteq P}}$.
On the other hand, considering arbitrary differences of support functions gives rise to \defn{virtual polyhedra}, which is the Grothendieck group of the semigroup of polytopes with Minkowski sums~\cite{PukhlikovKhovanskii}.

@@

\begin{theorem}
\label{thm:associahedronLoday}
The associahedron~$\Asso[n]$ can be described equivalently as:
\begin{itemize}
\item the convex hull of the points~$L(T) \in \R^{n+1}$ for all rooted binary trees~$T$ with~$n+1$ internal nodes, where the $i$th coordinate of~$L(T)$ is the product of the number of leaves in the left subtree by the number of leaves in the right subtree of the $i$th node of~$T$ in inorder~\cite{Loday},
\item the intersection of the hyperplane~$\bigset{\b{x} \in \R^{n+1}}{\sum_{\ell \in [n+1]} \b{x}_\ell = \binom{n+2}{2}}$ with the halfspaces $\bigset{\b{x} \in \R^{n+1}}{\sum_{a < \ell < b} \b{x}_\ell \ge \binom{b-a}{2}}$ for all internal diagonals~$(a,b)$ of \mbox{the $(n+3)$-gon~\cite{ShniderSternberg}},
\item the Minkowski sum of the faces~$\simplex_{[i,j]}$ of the standard $n$-dimensional simplex~$\simplex_{[n+1]}$ corresponding to all proper intervals~$[i,j]$ of~$[n+1]$,~\cite{Postnikov}.
\end{itemize}
\end{theorem}

\enlargethispage{.1cm}
We now focus on the normal fan~$\gvectorFan[n]$ of~$\Asso[n]$.
Since~$\Asso[n]$ lives in a hyperplane of~$\R^{n+1}$, so does its normal fan.
Let~$\HH \eqdef \bigset{\b{x} \in \R^{n+1}}{\sum_{\ell \in [n+1]} \b{x}_\ell = 0}$ and~$\pi : \R^{n+1} \to \HH$ denote the orthogonal projection.
The following fans are illustrated in \cref{fig:lodayFans} in dimension~$2$ and~$3$.

\begin{theorem}
\label{thm:normalFanLoday}
In the normal fan~$\gvectorFan[n]$ of~$\Asso[n]$, 
\begin{itemize}
\item the normal vector of the facet corresponding to an internal diagonal~$(a,b)$ of the $(n+3)$-gon is the vector~$\gvector{a,b} \eqdef \pi\big( \sum_{a < \ell < b} \b{e}_\ell \big) = (n+1) \sum_{a < \ell < b} \b{e}_\ell - (b-a-1) \sum_{1 \le \ell \le n+1} \b{e}_\ell$.
\item the normal cone of the vertex corresponding to a rooted binary tree~$T$ is the incidence cone~$\bigset{\b{x} \in \HH}{\b{x}_i \le \b{x}_j \text{ for all edges $i \to j$ in $T$}}$.
\end{itemize}
\end{theorem}

@@

\para{Principal coefficients and $\b{g}$- and $\b{c}$-vectors}
%
We now consider principal coefficients to define the $\b{g}$- and $\b{c}$-vectors.
As defined in~\cite[Def.~3.1]{FominZelevinsky-ClusterAlgebrasIV}, the cluster algebra with \defn{principal coefficients} at~$\B_\circ$ is the cluster algebra~$\clusterAlgebra[\B_\circ][\coefficients_\circ]$ in~$\Q(x_1, \dots, x_n, p_1, \dots, p_n)$, where~$\coefficients_\circ = \{p_x\}_{x \in \cluster_\circ}$ are precisely the generators~$p_1, \dots, p_n$ of~$\Trop{n}$ relabeled by~$\cluster_\circ$.
For simplicity, we drop the mention of the coefficients for principal cluster algebras in the notations for seeds~$\seed = (\B, \cluster)$, for cluster algebras~$\principalClusterAlgebra$ and for variables~$\principalVariables$.
Principal coefficients cluster algebras are $\Z^n$-graded (in the fundamental weight basis~$\{\simpleWeight_x\}_{x \in \cluster_\circ}$ of~$V$) for the degree function~$\deg(\B_\circ,\cdot)$ on $\principalClusterAlgebra$ obtained by setting
\(
\deg(\B_\circ,  x) \eqdef \fundamentalWeight_x
\)
and
\(
\deg(\B_\circ, p_x) \eqdef \sum_{y \in \cluster_\circ} -b_{yx} \fundamentalWeight_y
\)
for any $x\in\cluster_\circ$.

@@

%\newpage
\part{Grothendieck group and relations between $\b{g}$-vectors}
\label{part:algebra}

@@

%\newpage
\bibliographystyle{alpha}
\bibliography{typeConeAssociahedra}
\label{sec:biblio}
